---
title: "Darwin Core mapping"
subtitle: "For: Registry of non-native species in the Two Seas region countries (Great Britain, France, Belgium and the Netherlands)"
author:
- Lien Reyserhove
- Peter Desmet
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

This document describes how we map the checklist data to Darwin Core. The source file for this document can be found [here](https://github.com/trias-project/rinse-checklist/blob/master/data/raw/oo_30975.xlsx).

# Setup

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Set locale (so we use UTF-8 character encoding):

```{r}
# This works on Mac OS X, might not work on other OS
Sys.setlocale("LC_CTYPE", "en_US.UTF-8")
```

Load libraries:

```{r}
library(tidyverse) # To transform data
library(magrittr)  # For %<>% pipes
library(janitor)   # To clean input data
library(readxl)    # To read Excel files
library(digest)    # To generate hashes
library(rgbif)     # Interface to the GBIF API
```

Set file paths (all paths should be relative to this script):
 
```{r}
# Raw files:
raw_data_file = "../data/raw/oo_30975.xlsx"

# Processed files:

```

# Read and pre-process raw data

Create a data frame `raw_data` from the source data:

```{r}
# Read the source data:
raw_data <- read_excel(raw_data_file, sheet = "Registry")
```

Clean the data somewhat: remove empty rows if present

```{r}
raw_data %<>%
  remove_empty("rows") %>%     # Remove empty rows
  clean_names()               # Have sensible (lowercase) column names
```

## Generate taxonID

To uniquely identify a taxon in the taxon core and reference taxa in the extensions, we need a `taxonID`. Since we need it in all generated files, we generate it here in the raw data frame. It is a combination of `dataset-shortname:taxon:` and a hash based on the scientific name. As long as the scientific name doesn't change, the ID will be stable.

The scientific name should be the name in lowest level taxonomic rank that can be determined, which is the species name (genus + specificEpithet) here. This information is a compilation of `genus` and `species` in `raw_data`. To determine the quality of these fields, we  paste `genus` and `species` together in `scientifcName` and we parse them with the GBIF nameparser. This will give us an impression of the quality of the scientificName

Paste `genus` and `species` into `scientificName`:

```{r}
raw_data %<>% mutate(scientificName = paste(genus, species))
```

Parse names:

```{r}
parsed_names <- parsenames(raw_data $ scientificName)
```

Retain the columns of interest to scan the quality (i.e `scientificName`, `type`, `parsed` and `rankmarker`)

```{r}
parsed_names %<>% select(scientificname, type, parsed, rankmarker) 
```

ScientificName contains different types of taxa:

```{r, echo = FALSE}
parsed_names %>% select(type, parsed, rankmarker) %>% 
  group_by_all() %>% 
  summarize(records = n())
```

We merge this information with `raw_data`:

```{r}
raw_data %<>%  left_join(parsed_names, by = c("scientificName" = "scientificname"))
```

Information with respect to the hybrids and the informal types of taxa might need some cleaning (see [this issue](https://github.com/trias-project/rinse-checklist/issues/3) on GitHub. We inspect these values:

```{r}
raw_data %>% select(scientificName, type) %>% 
  filter(type == "INFORMAL" | type == "HYBRID") %>% 
  arrange(type)
```

Within the `INFORMAL` group, we clean information in `species` by removing the specificEpiphet `agg.` and `sp.:

```{r}
raw_data %<>% mutate(species_clean = case_when(
  species == "agg." ~ "",
  species == "sp." ~ "",
  TRUE ~ species))
```

Within the `HYBRID` group, some specificEpithet also need some cleaning:

```{r}
raw_data %<>% mutate(species_clean = case_when(
  scientificName == "Brachyglottis x 'Sunshine'" ~  "'Sunshine'",
  scientificName == "Cotoneaster x 'Hybridus pendulus'" ~ "'Hybridus pendulus'",
  scientificName == "Fuchsia x 'Corallina'" ~ "'Corallina'",
  scientificName == "Symphytum x 'Hidcote Blue'" ~ "'Hidcote Blue'"))
```

Based on the cleaning of the species information, we generate `scientificName_clean`:



```{r}
raw_data %>% select(scientificName_clean, type) %>% 
  filter(type == "INFORMAL" | type == "HYBRID") %>% 
  arrange(type)
```

Create taxonID:

```{r}
# Vectorize the digest function (The digest() function isn't vectorized. So if you pass in a vector, you get one value for the whole vector rather than a digest for each element of the vector):
vdigest <- Vectorize(digest)

# Generate taxonID:
raw_data %<>% mutate(taxon_id = paste("rinse-checklist", "taxon", vdigest(scientificName_clean, algo="md5"), sep = ":"))
```

## Further pre-processing:

Add prefix `raw_` to all column names to avoid name clashes with Darwin Core terms:

```{r}
colnames(raw_data) <- paste0("raw_", colnames(raw_data))
```

Preview data:

```{r}
raw_data %>% head()
```

# Create taxon core

```{r start_taxon}
taxon <- raw_data
```

## Term mapping
 
Map the data to [Darwin Core Taxon](http://rs.gbif.org/core/dwc_taxon_2015-04-24.xml).
 
```{r}
taxon %<>% mutate(language = "en")
```

### license

```{r}
taxon %<>% mutate(license = "http://creativecommons.org/publicdomain/zero/1.0/")
```

### rightsHolder

```{r}
taxon %<>% mutate(rightsHolder = "University of Cambridge")

```

### accessRights

```{r}
taxon %<>% mutate(accessRights = "http://www.inbo.be/en/norms-for-data-use")
```

### datasetID

```{r}
taxon %<>% mutate(datasetID = "")
```

### datasetName

```{r}
taxon %<>% mutate(datasetName = "Registry of non-native species in the Two Seas region countries (Great Britain, France, Belgium and the Netherlands")
```

### taxonID

```{r}
taxon %<>% mutate(taxonID = raw_taxon_id)
