---
title: "Darwin Core mapping"
subtitle: "For: Registry of non-native species in the Two Seas region countries (Great Britain, France, Belgium and the Netherlands)"
author:
- Lien Reyserhove
- Peter Desmet
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

This document describes how we map the checklist data to Darwin Core. The source file for this document can be found [here](https://github.com/trias-project/rinse-checklist/blob/master/data/raw/oo_30975.xlsx).

# Setup

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Set locale (so we use UTF-8 character encoding):

```{r}
# This works on Mac OS X, might not work on other OS
Sys.setlocale("LC_CTYPE", "en_US.UTF-8")
```

Load libraries:

```{r}
library(tidyverse) # To transform data
library(magrittr)  # For %<>% pipes
library(janitor)   # To clean input data
library(readxl)    # To read Excel files
library(digest)    # To generate hashes
library(rgbif)     # Interface to the GBIF API
```

Set file paths (all paths should be relative to this script):
 
```{r}
# Raw files:
raw_data_file = "../data/raw/oo_30975.xlsx"

# Processed files:

```

# Read and pre-process raw data

Create a data frame `raw_data` from the source data:

```{r}
# Read the source data:
raw_data <- read_excel(raw_data_file, sheet = "Registry")
```

Clean the data somewhat: remove empty rows if present

```{r}
raw_data %<>%
  remove_empty("rows") %>%     # Remove empty rows
  clean_names()               # Have sensible (lowercase) column names
```

## Clean scientific names

The scientific name should be the name in lowest level taxonomic rank that can be determined, which is the species name (genus + specificEpithet) here. This information is a compilation of `genus` and `species` in `raw_data`. Before we assign a taxonID to these scientific names, we need to clean this information. We do this stepwise:

1. Paste `genus` and `species` together in `scientifcName` 
2. Parse `scientificName` with the GBIF nameparser. This output of this function give us an impression of the quality of the generated `scientificName` (see further)
3. Screen quality of `scientificName` on predefined criteria
4. Generate `scientificName_clean` 
5. Generate `genus_clean`, `species_clean` and `infraspecificEpithet_clean` by the information provided by the nameparser function.

Paste `genus` and `species` into `scientificName`:

```{r}
raw_data %<>% mutate(scientificName = paste(genus, species))
```

Parse names:

```{r}
parsed_names <- parsenames(raw_data $ scientificName)
```

For the cleaning of the scientific names, we screen on the following criteria:
- `type`(should be scientific) OR
- `parsed` (should be TRUE)  OR
- `parsedpartially` (should be FALSE)  OR
- `authorship` (should be empty)

```{r}
parsed_names %>% mutate(authorship_empty = case_when(
  is.na(authorship) ~ "NA",
  !is.na(authorship) ~ "provided")) %>% 
  select(type, parsed, parsedpartially, authorship_empty) %>% 
  group_by_all() %>% 
  summarize(count = n())
```

All scientificNames for which `type = SCIENTIFIC` AND `PARSED = TRUE` AND `parsedpartially = FALSE` AND `authorship_empty = NA` will not be screened (6571 records) for cleaning. This is an overview of the remaining scientific names to be screened:

```{r}
parsed_names %>% select(scientificname, genusorabove, specificepithet, infraspecificepithet, type, parsed, parsedpartially, authorship, rankmarker) %>% 
  filter(!(type == "SCIENTIFIC" & parsed == "TRUE" & parsedpartially == "FALSE" & is.na(authorship))) %>% 
  arrange(type)
```

The decisions for the cleaning steps are summarized in [this issue](https://github.com/trias-project/rinse-checklist/issues/3)

For some scientific names, we need information on `authorship` to clean them, so we merge this information with the raw dataset:

```{r}
# Select the necessary columns in `parsed_names`:
parsed_names %<>% select(scientificname, authorship) 

# Merge with raw_data:
raw_data %<>% left_join(parsed_names, by = c("scientificName" = "scientificname"))

```

Generate `scientificName_clean`:

```{r}
raw_data %<>% mutate(scientificName_clean = case_when(
   scientificName == "Brachyglottis x 'Sunshine'" ~ "Brachyglottis 'Sunshine'",
   scientificName == "Cotoneaster x 'Hybridus pendulus'" ~ "Cotoneaster 'Hybridus pendulus'",
   scientificName == "Fuchsia x 'Corallina'" ~ "Fuchsia 'Corallina'",
   scientificName == "Symphytum x 'Hidcote Blue'" ~ "Symphytum 'Hidcote Blue'",
   species == "agg." ~ genus,
   species == "sp." ~ genus,
   authorship == "Orientalis" ~ "Anchusa arvensis orientalis",
   authorship == "Italicum" ~ "Arum italicum italicum",
   authorship == "Officinalis" ~ "Asparagus officinalis officinalis",
   authorship == "Oleifera" ~ "Brassica rapa oleifera",
   authorship == "Rapa" ~ "Brassica rapa rapa",
   authorship == "Inermis" ~ "Bromus inermis inermis",
   authorship == "Glandulosa" ~ "Calamintha nepeta glandulosa",
   authorship == "Carpatica" ~ "Anthyllis vulneraria carpatica",
   is.na(species) ~ genus,
   scientificName == "Rosa Hollandica'" ~ "Rosa 'Hollandica'",
   scientificName == "Sedum Herbstfreude'" ~ "Sedum 'Herbstfreude'",
   species == "x" ~ genus,
   scientificName == "Lipeurus maculosusIschnocera" ~ "Lipeurus maculosus",
   species == "n.a." ~ genus,
   scientificName == "AseroÙ rubra" ~ "Asero rubra",
   scientificName == "Cuprocyparis x leylandii hyb." ~ "Cuprocyparis x leylandii",
   scientificName == "Larix x marschlinsii hyb." ~ "Larix x marschlinsii",
   TRUE ~ scientificName))
 
```

We will now create `genus_clean`, `species_clean` and `infraspecificEpithet_clean` by applying the nameparser on `scientificName_clean`:

```{r}
parsed_names_clean <- parsenames(raw_data $ scientificName_clean)
```

Select required columns:

```{r}
parsed_names_clean %<>% select(scientificname, type, genusorabove, specificepithet, infraspecificepithet, rankmarker) 
```

Merge `parsed_names_clean` with `raw_data`:

```{r}
raw_data %<>% left_join(parsed_names_clean, by = c("scientificName_clean" = "scientificname")) 
```

Rename column names:

```{r}
raw_data %<>% rename(genus_clean = genusorabove,
                     specificEpithet_clean = specificepithet,
                     infraspecificEpithet_clean = infraspecificepithet) 
```

## Create taxonID:

To uniquely identify a taxon in the taxon core and reference taxa in the extensions, we need a `taxonID`. Since we need it in all generated files, we generate it here in the raw data frame. It is a combination of `dataset-shortname:taxon:` and a hash based on the scientific name. As long as the scientific name doesn't change, the ID will be stable.

```{r}
# Vectorize the digest function (The digest() function isn't vectorized. So if you pass in a vector, you get one value for the whole vector rather than a digest for each element of the vector):
vdigest <- Vectorize(digest)

# Generate taxonID:
raw_data %<>% mutate(taxon_id = paste("rinse-checklist", "taxon", vdigest(scientificName_clean, algo="md5"), sep = ":"))
```

## Further pre-processing:

Add prefix `raw_` to all column names to avoid name clashes with Darwin Core terms:

```{r}
colnames(raw_data) <- paste0("raw_", colnames(raw_data))
```

Preview data:

```{r}
raw_data %>% head()
```

# Create taxon core

```{r start_taxon}
taxon <- raw_data
```

## Term mapping
 
Map the data to [Darwin Core Taxon](http://rs.gbif.org/core/dwc_taxon_2015-04-24.xml).
 
```{r}
taxon %<>% mutate(language = "en")
```

### license

```{r}
taxon %<>% mutate(license = "http://creativecommons.org/publicdomain/zero/1.0/")
```

### rightsHolder

```{r}
taxon %<>% mutate(rightsHolder = "University of Cambridge")

```

### accessRights

```{r}
taxon %<>% mutate(accessRights = "http://www.inbo.be/en/norms-for-data-use")
```

### datasetID

```{r}
taxon %<>% mutate(datasetID = "")
```

### datasetName

```{r}
taxon %<>% mutate(datasetName = "Registry of non-native species in the Two Seas region countries (Great Britain, France, Belgium and the Netherlands")
```

### taxonID

```{r}
taxon %<>% mutate(taxonID = raw_taxon_id)
```

### scientificName

```{r}
taxon %<>% mutate(scientificName = raw_scientificName_clean)
```

### kingdom

Information on the kingdoms is not available in the raw data. However, we can derive it form the field `raw_phylum_division` using rgbif.

For this, we need to save all phyla in a separate dataframe: 

```{r}
phylum <- taxon %>% distinct(raw_phylum_division)
```

```{r, echo=FALSE}
(phylum)
```

We remove `Vira` from the list as this is there's no kingdom for this group.

```{r}
phylum %<>% filter(raw_phylum_division != "Vira") 
```

Select `kingdkom` using rgbif:

```{r}
phylum %<>% rowwise() %>%
 mutate(kingdom = name_usage(name = raw_phylum_division, 
                             return = "data")  %>% 
          select(kingdom) %>% 
          filter(!is.na(kingdom)) %>%
          count(kingdom) %>%
          filter(n == max(n)) %>%
          slice(1L) %>%
          pull(kingdom)) %>%
 ungroup()
```

Merge `phylum` with `taxon`:

```{r}
taxon %<>% left_join(phylum, by = "raw_phylum_division")
```

### phylum

```{r}
taxon %<>% mutate(phylum = raw_phylum_division) 
```

### class

```{r}
taxon %<>% mutate(class = raw_class) 
```

### genus

```{r}
taxon %<>% mutate(genus = raw_genus)
```


COMPLETE


# Create distribution extension

Map the data to [Species Distribution](http://rs.gbif.org/extension/gbif/1.0/distribution.xml).

## Pre-processing

```{r}
distribution <- raw_data
```

## Term mapping

### taxonID

```{r}
distribution %<>% mutate(taxonID = ) 
```

Information for `locationID`, `locality`, `countrycode` and `occurrenceStatus` can be found in the columns `raw_great_brittain`, `raw_france`, `raw_belgium` and `raw_netherlands`, which represent the occurrence for each taxon in each country. For further mapping, we need to integrate the country information in one column `country` and the occurrence information in a second column:

```{r}
distribution %<>% gather(country, occurrence, raw_great_britain, raw_france, raw_belgium, raw_netherlands, na.rm = TRUE) 
```

Clean country names:

```{r}
distribution %<>% mutate(country = recode(country,
    raw_great_britain  = "Great Britain",
    raw_france         = "France",
    raw_belgium        = "Belgium",
    raw_netherlands    = "The Netherlands"))
```


### locationID

```{r}
distribution %<>% mutate(locationID = case_when(
    country == "Great Britain"   ~ "ISO_3166-2:GB",
    country == "France"          ~ "ISO_3166-2:FR",
    country == "Belgium"         ~ "ISO_3166-2:BE",
    country == "The Netherlands" ~ "ISO_3166-2:NL"))
```

### locality

```{r}
distribution %<>% mutate(locality = country) 
```

### countryCode

```{r}
distribution %<>% mutate(countryCode = case_when(
    country == "Great Britain"   ~ "GB",
    country == "France"          ~ "FR",
    country == "Belgium"         ~ "BE",
    country == "The Netherlands" ~ "NL")) 
```

### occurrenceStatus

Information for `occurrenceStatus` is contained in `occurrence`, which contains the following unique values:

```{r}
distribution %>% distinct(occurrence)
```


